---
title: "Data Analysis"
author: "Emily Alsentzer and Julie Zhu"
date: "May 31, 2016"
output: html_document
---
GOAL #1
First, we will perform a chi squared test of homogenity to see whether or not the distribution of race is the same for each type of "armed" status. Blacks make up 35% of all unarmed police killings, even though they make up 6% of the entire population.
```{r setup, include=FALSE}
library(readr)
police_killings <- read.csv("data/police_killings.csv")
armed_v_race <- data.frame(as.character(police_killings$race),  as.character(police_killings$armed))
contingency_table <- t(table(armed_v_race))[,c(2,3,6)]
chifit <- chisq.test(contingency_table)
(chifit$residuals)^2

unarmed <- contingency_table[4,]
black_unarmed <- unlist(unname(unarmed[2]))/sum(unarmed)
white_unarmed <- unlist(unname(unarmed[6]))/sum(unarmed)
```

We can see that the largest contribution comes from not armed and armed with firearm. We remove and retest. 
```{r}
#remove armed= "No"
ct2 <- contingency_table[-4,]
chisq.test(ct2)
```

The result is still significant so we also remove firearm. 
```{r}
#remove armed= "Firearm"
ct3 <- ct2[-2,]
chisq.test(ct3)
```
Now the test no longer yields a significant result. 

GOAL 2
The goal is to see whether there is a difference in the distribution of races between killed population and normal population.
```{r}
library(plyr)
race_summary <- table(police_killings$race)
true_pop <- total_race[,2]#/ sum(total_race[,2])*100000
killed_rop <- race_summary#/sum(race_summary)*100000
both_populations <- cbind(unlist(unname(true_prop)), unname(killed_prop))
colnames(both_populations) <- c("true_prop", "killed_prop")
both_populations_t <- t(both_populations)
```

Now we will perform a chi squared test of homogenity to see whether or not the distribution of race is the same for the killed population and normal population.

```{r}
chifit <- chisq.test(both_populations_t)
(chifit$residuals)^2

chifit2 <- chisq.test(both_populations)
(chifit2$residuals)^2
```

GOAL 3
RELATIVE RISK
```{r}
library(abd)
library(epitools)
blacks_killed <- police_killings[which(police_killings$race == "Black"),]
whites_killed <- police_killings[which(police_killings$race == "White"),]
histpanics_killed <- police_killings[which(police_killings$race == "Hispanic/Latino"),]


total_race = read.csv("clean_data/total_race.csv")
killed <- c(nrow(blacks_killed), nrow(whites_killed))
not_killed <- c(total_race[2,2] - nrow(blacks_killed), total_race[6,2] - nrow(whites_killed))
twoBytwo <- c(killed[1], not_killed[1], killed[2], not_killed[2])
rownames(twoBytwo) <- c("Black", "White")
oddsratio(twoBytwo, method="wald", conf.level=0.95)


relative_risk <- function(race1_killed, race2_killed, total_race1, total_race2){
    black_p <- race1_killed/total_race1
    white_p <- race2_killed/total_race2
    RR <- black_p/white_p  
    return(RR)
}

RR_CI <- function(rr, x1, x2, n1, n2){
    SE <- sqrt( ((n1-x1)/x1)/n1 + ((n2-x2)/x2)/n2 )
    lower <- log(rr) - 1.96 * SE
    upper <- log(rr) + 1.96 * SE
    return (c(exp(lower),exp(upper)))
}

# x1 = nrow(blacks_killed)
# x2 =  nrow(whites_killed)
# n1 = total_race[2,2]
# n2 = total_race[6,2])

rr_bw <- relative_risk(nrow(blacks_killed), nrow(whites_killed), total_race[2,2], total_race[6,2])
RR_CI(rr_bw, nrow(blacks_killed),nrow(whites_killed), total_race[2,2], total_race[6,2])

rr_hw <- relative_risk(nrow(histpanics_killed), nrow(whites_killed), total_race[3,2], total_race[6,2])
RR_CI(rr_hw, nrow(histpanics_killed), nrow(whites_killed), total_race[3,2], total_race[6,2])

rr_bh <- relative_risk(nrow(blacks_killed), nrow(histpanics_killed),  total_race[2,2], total_race[3,2])
RR_CI(rr_bh, nrow(blacks_killed),nrow(histpanics_killed), total_race[2,2], total_race[3,2])



```

Relative risk for armed vs not armed
```{r}
not_armed <- police_killings[which(police_killings$armed == "No"),]
armed <- police_killings[which(police_killings$armed != "No"),]

unarmed_blacks_killed <- not_armed[which(not_armed$race == "Black"),]
unarmed_whites_killed <- not_armed[which(not_armed$race == "White"),]
unarmed_histpanics_killed <- not_armed[which(not_armed$race == "Hispanic/Latino"),]

rr_bw <- relative_risk(nrow(unarmed_blacks_killed), nrow(unarmed_whites_killed), total_race[2,2], total_race[6,2])
RR_CI(rr_bw, nrow(unarmed_blacks_killed),nrow(unarmed_whites_killed), total_race[2,2], total_race[6,2])

rr_hw <- relative_risk(nrow(unarmed_histpanics_killed), nrow(unarmed_whites_killed), total_race[3,2], total_race[6,2])
RR_CI(rr_hw, nrow(unarmed_histpanics_killed), nrow(unarmed_whites_killed), total_race[3,2], total_race[6,2])

rr_bh <- relative_risk(nrow(unarmed_blacks_killed), nrow(unarmed_histpanics_killed),  total_race[2,2], total_race[3,2])
RR_CI(rr_bh, nrow(unarmed_blacks_killed),nrow(unarmed_histpanics_killed), total_race[2,2], total_race[3,2])

armed_blacks_killed <- armed[which(armed$race == "Black"),]
armed_whites_killed <- armed[which(armed$race == "White"),]
armed_histpanics_killed <- armed[which(armed$race == "Hispanic/Latino"),]

rr_bw <- relative_risk(nrow(armed_blacks_killed), nrow(armed_whites_killed), total_race[2,2], total_race[6,2])
RR_CI(rr_bw, nrow(armed_blacks_killed),nrow(armed_whites_killed), total_race[2,2], total_race[6,2])

rr_hw <- relative_risk(nrow(armed_histpanics_killed), nrow(armed_whites_killed), total_race[3,2], total_race[6,2])
RR_CI(rr_hw, nrow(armed_histpanics_killed), nrow(armed_whites_killed), total_race[3,2], total_race[6,2])

rr_bh <- relative_risk(nrow(armed_blacks_killed), nrow(armed_histpanics_killed),  total_race[2,2], total_race[3,2])
RR_CI(rr_bh, nrow(armed_blacks_killed),nrow(armed_histpanics_killed), total_race[2,2], total_race[3,2])


```



GOAL 4
Next we want to investigate what the socioeconomic status is between the areas where people of different races were killed.
compare poverty, education, income, etc between areas where blacks vs whites were killed

```{r}
blacks_killed <- police_killings[which(police_killings$race == "Black"),]
whites_killed <- police_killings[which(police_killings$race == "White"),]
hispanics_killed <- police_killings[which(police_killings$race == "Hispanic/Latino"),]
native_american_killed <- police_killings[which(police_killings$race == "Native American"),]


census_test <- function(blacks, whites, race1, race2){
    wilcox <- wilcox.test(blacks, whites)
   # t.test(blacks, whites) 
    black_mean = mean(blacks)
    white_mean =  mean(whites)
    census <-data.frame(factor(c(race1,  race2)), c(blacks, whites))
    colnames(census) <- c("Race", "Indicator")

  #  ggplot(census, aes(x=Indicator, fill=Race)) +
    #    geom_histogram( alpha=.5, position="identity")

    #2 = black; 6 = white
    boxplot(Indicator~Race, data=census)
    return (c(wilcox, black_mean, white_mean))
}

#INCOME
bw_income <- census_test(blacks_killed$med_income, whites_killed$med_income, blacks_killed$race, whites_killed$race) #significant
hw_income <- census_test(hispanics_killed$med_income, whites_killed$med_income, hispanics_killed$race, whites_killed$race) #significant
hb_income <- census_test(hispanics_killed$med_income, blacks_killed$med_income, hispanics_killed$race, blacks_killed$race) #significant

#POVERTY
bw_poverty <- census_test(blacks_killed$poverty_pct, whites_killed$poverty_pct, blacks_killed$race, whites_killed$race) # Not significant
bh_poverty <- census_test(blacks_killed$poverty_pct, hispanics_killed$poverty_pct, blacks_killed$race, hispanics_killed$race)
hw_poverty <- census_test(whites_killed$poverty_pct, hispanics_killed$poverty_pct, whites_killed$race, hispanics_killed$race)

#EDUCATION
bw_educ <- census_test(blacks_killed$educ_pct, whites_killed$educ_pct, blacks_killed$race, whites_killed$race) # Not significant
hw_educ <- census_test(hispanics_killed$educ_pct, whites_killed$educ_pct, hispanics_killed$race, whites_killed$race) # Not significant
bh_educ <- census_test(hispanics_killed$educ_pct, blacks_killed$educ_pct, hispanics_killed$race, blacks_killed$race) # Not significant


p_values <- c(bw_income$p.value, hw_income$p.value,hb_income$p.value, bw_poverty$p.value, bh_poverty$p.value, hw_poverty$p.value, bw_educ$p.value, hw_educ$p.value, bh_educ$p.value)
p.adjust(p_values, method = "holm", n = length(p_values)) # holm method

```

Now we will compare the values for the killed population to the national averages.
```{r}
poverty_dataset <- read_csv("clean_data/poverty_income.csv", col_names = TRUE)
wilcox <- wilcox.test(poverty_dataset$med_income, police_killings$med_income)
wilcox <- wilcox.test(poverty_dataset$med_income, blacks_killed$med_income)

poverty_dataset <- data_frame(race = rep("U.S.", nrow(poverty_dataset)),
                              med_income = poverty_dataset$med_income,
                              poverty_pct = poverty_dataset$poverty_pct)

police_killings %>%
  select(race, med_income, poverty_pct) %>%
  bind_rows(poverty_dataset) %>%
  mutate(race = factor(race, levels = c("U.S.", "Black", "White", 
                                        "Hispanic/Latino"))) %>%
  filter(race %in% c("U.S.", "Black", "White", "Hispanic/Latino")) %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = race, y = med_income))

boxplot(cbind(poverty_dataset$med_income, blacks_killed$med_income, whites_killed$med_income, hispanics_killed$med_income), names=c("U.S.", "Black", "White", "Hispanic/Latino"), xlab = "Population", ylab= "Median Household Income")

boxplot(cbind(poverty_dataset$poverty_pct, blacks_killed$poverty_pct, whites_killed$poverty_pct, hispanics_killed$poverty_pct), names=c("U.S.", "Black", "White", "Hispanic/Latino"), xlab = "Population", ylab= "Percent Below the Poverty Line")

education <- read_csv("clean_data/education_attainment.csv", col_names = TRUE)
colnames(education) <- c("state", "county", "educ_pct")
boxplot(cbind(education$educ_pct, blacks_killed$educ_pct, whites_killed$educ_pct, hispanics_killed$educ_pct), names=c("U.S.", "Black", "White", "Hispanic/Latino"), xlab = "Population", ylab= "Percent with High School Diploma/ GED", title ="title")
```


