---
title: "Data Analysis"
author: "Emily Alsentzer and Julie Zhu"
date: "May 31, 2016"
output: html_document
---
GOAL #1
First, we will perform a chi squared test of homogenity to see whether or not the distribution of race is the same for each type of "armed" status. Blacks make up 35% of all unarmed police killings, even though they make up 6% of the entire population.
```{r setup, include=FALSE}
police_killings <- read.csv("data/police_killings.csv")
armed_v_race <- data.frame(as.character(police_killings$race),  as.character(police_killings$armed))
contingency_table <- t(table(armed_v_race))
chifit <- chisq.test(contingency_table)
(chifit$residuals)^2

unarmed <- contingency_table[4,]
black_unarmed <- unlist(unname(unarmed[2]))/sum(unarmed)
white_unarmed <- unlist(unname(unarmed[6]))/sum(unarmed)
```

We can see that the largest contribution comes from not armed and armed with firearm. We remove and retest. 
```{r}
#remove armed= "No"
ct2 <- contingency_table[-4,]
chisq.test(ct2)
```

The result is still significant so we also remove firearm. 
```{r}
#remove armed= "Firearm"
ct3 <- ct2[-2,]
chisq.test(ct3)
```
Now the test no longer yields a significant result. 

GOAL 2
The goal is to see whether there is a difference in the distribution of races between killed population and normal population.
```{r}
library(plyr)
race_summary <- table(police_killings$race)
true_pop <- total_race[,2]#/ sum(total_race[,2])*100000
killed_rop <- race_summary#/sum(race_summary)*100000
both_populations <- cbind(unlist(unname(true_prop)), unname(killed_prop))
colnames(both_populations) <- c("true_prop", "killed_prop")
both_populations_t <- t(both_populations)
```

Now we will perform a chi squared test of homogenity to see whether or not the distribution of race is the same for the killed population and normal population.
```{r}
chifit <- chisq.test(both_populations_t)
(chifit$residuals)^2

chifit2 <- chisq.test(both_populations)
(chifit2$residuals)^2
```

```{r}


```

