---
title: 'STATS 205: Graphs and Plots'
author: "Julie Zhu and Emily Alsentzer"
date: "May 31, 2016"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# Libraries
library(ggplot2)
library(purrr)
library(readr)
library(lubridate)
library(tidyr)
library(dplyr)
library(stringr)
library(datasets)
library(maps)
library(ca)
library(gridExtra)

# Parameters
file_in <- "data/police_killings.csv"
race_file <- "clean_data/county_race.csv"
total_file <- "clean_data/total_race.csv"
population_file <- "clean_data/state_pop.csv"
```

We first read in the data files. 

```{r}
df <- read_csv(file_in)
population <- read_csv(population_file, col_names = TRUE)
ethnicity <- read_csv(race_file, col_names = TRUE)
total_race <- read_csv(total_file, col_names = TRUE)
```

# Correspondence Analysis

## Armed vs. Race

```{r}
armed_race_ca <- table(df$armed, df$race) %>% as.matrix()
armed_ca <- ca(armed_race_ca)
summary(armed_ca)
plot(armed_ca, dim = c(1, 2), mass = c(TRUE, TRUE), 
     contrib = "absolute", map = "rowgreen", 
     arrows = c(FALSE, TRUE))
```

## Classification vs. Race

```{r}
class_race_ca <- table(df$classification, df$race)
class_ca <- ca(class_race_ca)
summary(class_ca)
plot(class_ca, dim = c(1, 2), mass = c(TRUE, TRUE), 
     contrib = "absolute", map = "rowgreen", 
     arrows = c(FALSE, TRUE))
```

# Environmental Circumstances Boxplots

## Unemployment Rate

```{r}
df %>%
  filter(!is.na(race), !is.na(unemployment_rate),
         race %in% c("White", "Black", "Hispanic/Latino")) %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = race, y = unemployment_rate)) +
  theme_bw() + 
  labs(x = "Ethnicity", y = "Unemployment Rate (%)")
```

## Educational Attainment

```{r}
df %>%
    filter(!is.na(race), !is.na(educ_pct),
         race %in% c("White", "Black", "Hispanic/Latino")) %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = race, y = educ_pct)) + 
  theme_bw() + 
  labs(x = "Ethnicity", y = "Educational Attainment (%)")
```

## Median Income

```{r}
df %>%
  filter(!is.na(race), !is.na(med_income),
         race %in% c("White", "Black", "Hispanic/Latino")) %>%  
  ggplot() +
  geom_boxplot(mapping = aes(x = race, y = med_income)) + 
  theme_bw() + 
  labs(x = "Ethnicity", y = "Median Income ($)")
```

## Poverty

```{r}
df %>%
  filter(!is.na(race), !is.na(poverty_pct),
         race %in% c("White", "Black", "Hispanic/Latino")) %>%  
  ggplot() +
  geom_boxplot(mapping = aes(x = race, y = poverty_pct)) +
  theme_bw() + 
  labs(x = "Ethnicity", y = "Poverty Rate (%)")
```

# Heatmaps

## Normalized State Killings

```{r, fig.width = 8, fig.height = 4}
states <- map_data("state")

new_df <- df %>%
  count(state, state_name) %>%
  left_join(population, by = c("state" = "state_abb")) %>%
  mutate(Count = n/pop)
  
map_df <- states %>%
  as.data.frame() %>%
  tbl_df() %>%
  select(-subregion) %>%
  left_join(new_df, by = c("region" = "state_name"))

map_df <- map_df[order(map_df$order),]

figure_1a <- map_df %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(mapping = aes(fill = Count)) +
  geom_path() + 
  theme_classic() + 
  scale_x_continuous(breaks = seq(-120, -70, by = 10)) +
  scale_fill_continuous(labels = scales::percent) +
  theme(axis.line.x = element_line(),
        axis.line.y = element_line()) +
  labs(x = "Longitude", y = "Latitude")
```

## Un-normalized State Killings

```{r, fig.width = 8, fig.height = 4}
new_df <- df %>%
  count(state_name) %>%
  rename(Count = n)

map_df <- states %>%
  as.data.frame() %>%
  tbl_df() %>%
  select(-subregion) %>%
  left_join(new_df, by = c("region" = "state_name"))

map_df <- map_df[order(map_df$order),]

map_df %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(mapping = aes(fill = Count)) +
  geom_path() + 
  theme_classic() + 
  scale_x_continuous(breaks = seq(-120, -70, by = 10)) +
  scale_fill_continuous(labels = scales::comma) +
  theme(axis.line.x = element_line(),
        axis.line.y = element_line()) +
  labs(x = "Longitude", y = "Latitude")
```

## Un-Normalized County Killings

```{r, fig.width = 8, fig.height = 4}
counties <- map_data("county")

new_df <- df %>%
  count(state_name, county) %>%
  rename(Count = n) %>%
  mutate(county = lapply(str_split(county, " "), `[[`, 1) %>% unlist())

map_df <- counties %>%
  as.data.frame() %>%
  tbl_df() %>%
  left_join(new_df, by = c("subregion" = "county", "region" = "state_name")) %>%
  filter(!is.na(Count))

map_df <- map_df[order(map_df$order),]

map_df %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(mapping = aes(fill = Count)) +
  geom_path(data = states, colour = "black") +
  scale_x_continuous(breaks = seq(-130, -70, by = 10)) +
  theme_classic() + 
  theme(axis.line.x = element_line(),
        axis.line.y = element_line()) +
  labs(x = "Longitude", y = "Latitude")
```

## Normalized County Killings

```{r, fig.width = 8, fig.height = 4}
new_df <- df %>%
  count(state_name, county) %>%
  left_join(ethnicity, by = c("county" = "county",
                              "state_name" = "state")) %>%
  filter(!is.na(total_county_pop)) %>%
  select(state_name, county, n, total_county_pop) %>%
  mutate(Count = n/total_county_pop,
         county = lapply(str_split(county, " "), `[[`, 1) %>% unlist())

map_df <- counties %>%
  as.data.frame() %>%
  tbl_df() %>%
  left_join(new_df, by = c("subregion" = "county", "region" = "state_name")) %>%
  filter(!is.na(Count))

map_df <- map_df[order(map_df$order),]

map_df %>%
  ggplot(mapping = aes(x = long, y = lat, group = group)) +
  geom_polygon(mapping = aes(fill = Count)) +
  geom_path(data = states, colour = "black") +
  theme_classic() + 
  theme(axis.line.x = element_line(),
        axis.line.y = element_line()) +
  scale_x_continuous(breaks = seq(-130, -70, by = 10)) +
  scale_fill_continuous(labels = scales::percent) +
  labs(x = "Longitude", y = "Latitude")
```

# Descriptive Graphs

## Age

```{r, fig.width = 4, fig.height = 4}
figure_2a <- df %>%
  filter(!is.na(age)) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = age), binwidth = 3) +
  theme_bw() + 
  labs(x = "Age", y = "Count")
```

## Gender

```{r, fig.width = 4, fig.height = 4}
figure_2b <- df %>%
  filter(!is.na(gender)) %>%
  ggplot() +
  geom_bar(mapping = aes(x = gender)) +
  theme_bw() + 
  labs(x = "Gender", y = "Count")
```

## Ethnicity

```{r, fig.width = 4, fig.height = 4}
figure_4a <- df %>%
  filter(!is.na(race)) %>%
  count(race) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(race, -n), y = n), 
           stat = "identity") +
  labs(x = "Ethnicity", y = "Count") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Normalized Ethnicity

```{r, fig.width = 5, fig.height = 4}
figure_4b <- df %>%
  count(race) %>%
  filter(!is.na(race)) %>%
  left_join(total_race, by = c("race" = "race")) %>%
  mutate(prop = (n/pop)*100000) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(race, -prop), y = prop), 
           stat = "identity") +
  labs(x = "Ethnicity", y = "Number Killed per 100,000") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 35, hjust = 1)) 
```

## Armed

```{r, fig.width = 5, fig.height = 4}
figure_3a <- df %>%
  count(armed) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(armed, -n), y = n), 
           stat = "identity") +
  labs(x = "Armed", y = "Count") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## Signs of Mental Illness

```{r, fig.width = 5, fig.height = 4}
figure_3b <- df %>%
  count(signs_of_mental_illness) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(signs_of_mental_illness, -n), y = n), 
           stat = "identity") +
  labs(x = "Signs of Mental Illness", y = "Count") +
  theme_bw() 
```

## Classification 

```{r, fig.width = 5, fig.height = 4}
figure_3c <- df %>%
  count(classification) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(classification, -n), y = n), 
           stat = "identity") +
  labs(x = "Manner of Death", y = "Count") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## State

```{r, fig.width = 5, fig.height = 4}
df %>%
  count(state) %>%
  top_n(10) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(state, -n), y = n), 
           stat = "identity") +
  labs(x = "State", y = "Count") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Normalized by Population State

```{r, fig.width = 5, fig.height = 4}
df %>%
  count(state) %>%
  top_n(10) %>%
  left_join(population, by = c("state" = "state_abb")) %>%
  filter(!(state == "DC")) %>%
  mutate(pop_pct = n/pop*100000) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(state, -pop_pct), y = pop_pct), 
           stat = "identity") +
  labs(x = "State", y = "Number Killed per 100,000") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Month

```{r, fig.width = 6, fig.height = 4}
figure_1b <- df %>%
  filter(!is.na(month(date))) %>%
  mutate(month = month(date)) %>%
  count(month) %>%
  ggplot() +
  geom_line(mapping = aes(x = factor(month), y = n), stat = "identity",
            group = 1) +
  labs(x = "Month", y = "Count") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_x_discrete(labels = c("January","February","March","April", "May", 
                              "June", "July", "August", "September",
                              "October", "November", "December"))
```

# Creating Figures

## Figure 1

```{r}
figure1 <- grid.arrange(figure_1a, figure_1b, ncol = 2, 
                        nrow = 1, widths = c(0.65, 0.35))
ggsave(plot = figure1, filename = "graphs/figure1.png",
       width = 11, height = 4)
```

## Figure 2

```{r}
figure2 <- grid.arrange(figure_2a, figure_2b, ncol = 2, nrow = 1, 
                        widths = c(0.5, 0.5))
ggsave(plot = figure2, filename = "graphs/figure2.png",
       width = 7, height = 4)
```

## Figure 3

```{r}
figure3 <- grid.arrange(figure_3a, figure_3b, figure_3c, ncol = 3, nrow = 1,
                        widths = c(0.4, 0.3, 0.4))
ggsave(plot = figure3, filename = "graphs/figure3.png",
       width = 11, height = 4)
```

## Figure 4

```{r}
figure4 <- grid.arrange(figure_4a, figure_4b, ncol = 2, nrow = 1,
                        widths = c(0.5, 0.5))
ggsave(plot = figure4, filename = "graphs/figure4.png",
       width = 7, height = 4)
```
