---
title: 'STATS 205: Pairwise Chi-squared and Binomial Tests'
author: "Julie Zhu"
date: "June 3, 2016"
output:
  html_document:
    toc: true
    number_sections: true
    theme: spacelab
    tango: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# Libraries
library(ggplot2)
library(purrr)
library(readr)
library(lubridate)
library(tidyr)
library(dplyr)
library(stringr)
library(datasets)

# Parameters
file_in <- "data/police_killings.csv"
race_file <- "clean_data/total_race.csv"
```

First, we read in the data. 

```{r}
df <- read_csv(file_in, col_names = TRUE)
total_race <- read_csv(race_file, col_names = TRUE)
```

# Chi-squared Test of Homogeneity

For all of the factors below, we want to compare whether the distribution
in factor level is the same across ethnicities. This means in our contigency 
table, ethnicities will be the columns and factors will be the rows. 

We limit to only three ethnicities due to the small cell adjustment. 

## Signs of Mental Illness

```{r}
mental_v_race <- df %>%
  select(race, signs_of_mental_illness) %>%
  filter(race %in% c("Black", "Hispanic/Latino", "White"))

contingency_table <- t(table(mental_v_race))
mentalfit <- chisq.test(contingency_table)

(mentalfit$residuals)^2
```

## Classification

```{r}
classification_v_race <- df %>%
  select(race, classification) %>%
  filter(race %in% c("Black", "Hispanic/Latino", "White"),
         classification != "Other")

contingency_table <- t(table(classification_v_race))
classificationfit <- chisq.test(contingency_table)

(classificationfit$residuals)^2
```

# Body Camera

```{r}
camera_v_race <- df %>%
  select(race, body_camera) %>%
  filter(race %in% c("Black", "Hispanic/Latino", "White"))

contingency_table <- t(table(camera_v_race))
camerafit <- chisq.test(contingency_table)

(camerafit$residuals)^2
```

## Threat Level

```{r}
threat_v_race <- df %>%
  select(race, threat_level) %>%
  filter(race %in% c("Black", "Hispanic/Latino", "White"))

contingency_table <- t(table(threat_v_race))
threatfit <- chisq.test(contingency_table)

(threatfit$residuals)^2
```

## Flee

```{r}
flee_v_race <- df %>%
  select(race, flee) %>%
  filter(race %in% c("Black", "Hispanic/Latino", "White"))

contingency_table <- t(table(flee_v_race))
fleefit <- chisq.test(contingency_table)

(fleefit$residuals)^2
```

## Armed

```{r}
armed_v_race <- df %>%
  select(race, armed) %>%
  filter(race %in% c("Black", "Hispanic/Latino", "White"))

contingency_table <- t(table(armed_v_race))
armedfit <- chisq.test(contingency_table)

(armedfit$residuals)^2
```

## Using the Holm Correction

```{r}
p_values <- c(armedfit$p.value, mentalfit$p.value, classificationfit$p.value,
              camerafit$p.value)

# Armed and Mental Illness are significant
p.adjust(p_values, method = "holm", n = 4) 
```

# Pair-wise Individual Ethnicity Tests

For each of the binomial tests, we see that there are two events: the victim
killed is of the race we are testing for or they are not. We will assume that
the probability of the victim being a certain race is determined by the 
proportion of that race in the United States population. 

```{r}
count_df <- df %>%
  filter(!is.na(race)) %>%
  count(race) 

n <- nrow(df)

us_pop <- sum(total_race$pop)
true_race_prop <- total_race %>%
  mutate(true_prop = pop/us_pop)
```

## Asian and Pacific Islander

```{r}
binom.test(count_df$n[1], n, p = true_race_prop$true_prop[1],
           alternative = "greater")

prop.test(count_df$n[1], n, p = true_race_prop$true_prop[1],
           alternative = "greater")

binom.test(count_df$n[1], n, p = true_race_prop$true_prop[1],
           alternative = "less")

prop.test(count_df$n[1], n, p = true_race_prop$true_prop[1],
           alternative = "less")
```

## Black

```{r}
binom.test(count_df$n[2], n, p = true_race_prop$true_prop[2],
           alternative = "greater")

prop.test(count_df$n[2], n, p = true_race_prop$true_prop[2],
           alternative = "greater")

binom.test(count_df$n[2], n, p = true_race_prop$true_prop[2],
           alternative = "less")

prop.test(count_df$n[2], n, p = true_race_prop$true_prop[2],
           alternative = "less")
```

## Hispanic/Latino

```{r}
binom.test(count_df$n[3], n, p = true_race_prop$true_prop[3],
           alternative = "greater")

prop.test(count_df$n[3], n, p = true_race_prop$true_prop[3],
           alternative = "greater")

binom.test(count_df$n[3], n, p = true_race_prop$true_prop[3],
           alternative = "less")

prop.test(count_df$n[3], n, p = true_race_prop$true_prop[3],
           alternative = "less")
```

## Native American

```{r}
binom.test(count_df$n[4], n, p = true_race_prop$true_prop[4],
           alternative = "greater")

prop.test(count_df$n[4], n, p = true_race_prop$true_prop[4],
           alternative = "greater")

binom.test(count_df$n[4], n, p = true_race_prop$true_prop[4],
           alternative = "less")

prop.test(count_df$n[4], n, p = true_race_prop$true_prop[4],
           alternative = "less")
```

## Other

```{r}
binom.test(count_df$n[5], n, p = true_race_prop$true_prop[5],
           alternative = "greater")

prop.test(count_df$n[5], n, p = true_race_prop$true_prop[5],
           alternative = "greater")

binom.test(count_df$n[5], n, p = true_race_prop$true_prop[5],
           alternative = "less")

prop.test(count_df$n[5], n, p = true_race_prop$true_prop[5],
           alternative = "less")
```

## White

```{r}
binom.test(count_df$n[6], n, p = true_race_prop$true_prop[6],
           alternative = "greater")

prop.test(count_df$n[6], n, p = true_race_prop$true_prop[6],
           alternative = "greater")

binom.test(count_df$n[6], n, p = true_race_prop$true_prop[6],
           alternative = "less")

prop.test(count_df$n[6], n, p = true_race_prop$true_prop[6],
           alternative = "less")
```
